The thin blue line:\\what protects my Android device?

### 2015-03-31 securityupdates.tex
Note that if build numbers do not change for updated code this indicates the manufacturer is incompetent

\section{Introduction}
We show that the security of Android relies on the centralisation of app deployment with Google.
Computer systems which run untrusted code are secure if there are strong sandboxing mechanisms which prevent the untrusted code doing dangerous things.
Google claims 100\% of Android devices have sandboxes and permissions~\cite{Patterson2013} but we show that on average only \daMeanSecurityPercTwosfNominal\ of Android devices have \emph{working} sandboxes and permissions due to known critical vulnerabilities.
The recurrent advice on keeping computers secure is to install the latest updates, yet only \daUpdatednessPercTwosfNominal\ of Android devices run the latest version and the rate of critical vulnerability discovery of between \avoVulnsPerYearAllAndroidTwosfNominal\ and \avoVulnsPerYearTwosfNominal\ per year is higher than the rate of updates being delivered of \daUpdatesPerYearTwosfNominal.
Yet Android malware deployments are not widespread -- despite the many attempts of malware authors:
in 2012, between 36.7\%~\cite{Zhou2012b} and 40\%~\cite{Zhou2012a} of malware for Android contained exploits designed to enable a malicious app to break the sandbox and thousands of new malware samples are found each year\todolater{citation needed}.

Android is protected by many layers of security and while the sandbox is usually broken on most Android devices by known vulnerabilities, the outer layers of protection such as the Google Play store prevent the installation of malicious apps.
The Google Play store provides economic (app developers have to pay to register), technical (Google Bouncer which scans apps) and social (user reviews and reports of malicious software) measures to prevent malicious apps being distributed.
Even users who cannot use Google Play (e.g.\ because they are in China) may still get its `Verify apps' functionality which scans for malicious apps.

The claim of the open source software `bazzar' is that it is more secure because anyone can review the source code and provide fixes~\cite{Raymond1999}.
Yet Android is not a `bazzar' in the way that Debian is because while much of the code is open source and available for the many eyes review, end users cannot easily rebuild the binaries that run on their phones due to the binary blobs and closed source components within them.
Since only manufacturers can do that, rather than any one technically competent user, the availability of updates is controlled by the commercial interests of the manufacturers (and the operators).
Manufacturers do not even provide updates for the length of the contracts under which the users originally bought the phone, despite efforts by the ACLU to persuade the FTC to force them to do so~\cite{Soghoian2013}, and despite the FTC forcing HTC to do so~\cite{FTCHTC2013}.
Since only manufacturers can provide updates, malicious parties cannot do so, if manufacturers promptly provided updates this would improve security, but since they do not, it makes it worse.

However, despite the current low level of malicious software deployment on Android, the high level of vulnerability and the slow deployment of updates is still a cause for concern.
Some users may only be able to use third party app stores or have USB debugging enabled and connect their phone to a malicious machine.
There have also been two remote code execution vulnerabilities in Android, one in the Javascript-to-Java bridge and one in dhcpd which allow a network attacker to run malicious code in a user context, bypassing all the checks on apps and leaving only the sandbox, which we show to be vulnerable.

Not all malicious apps need to break the sandbox in order to misbehave.
We focus on the issue of apps breaking out of the sandbox because malware which is able to take full control of a handset can do significantly more harm and is much harder to remove.
Ordinary malware can be uninstalled by the user, or remotely by Google through the Play Store, and returns the device to a secure state. 
Removing an app which has used an exploit to take control of the handset is unlikely to return the device to secure state.
We describe our threat model and three attack vectors in detail in \S\ref{sec:threatmodel}.

In order to determine whether an individual smartphone is exposed to a particular vulnerability, we need to combine vulnerability data with the operating system version string and build number from the handset.
For this we use handset data gathered from \daNumOSDevices\ devices in the \da~\cite{Wagner2013} project.
From this we have determined that since \daStartDate\ on five occasions all Android devices have been exposed to known critical vulnerabilities and that it takes a year for updates to reach most devices, by which time new vulnerabilities have been discovered.

In summary, the contributions of this paper are:
\begin{itemize}
 \item We show that the security of Android relies on the centralisation of app deployment with Google rather than the absence of vulnerabilities in its sandbox
 \item We quantify the Android update ecosystem showing how the OS version distribution changes and the frequency of updates.
 \item We quantify vulnerabilities by determining what proportion of devices they affected and how that changes over time.
 \item We created an open database to collate quantitative information on vulnerabilities which affect Android, and investigate their time-lines.
\end{itemize}

\section{Threat model}
\label{sec:threatmodel}

In this paper we measure whether an attacker can take control of a handset by breaking out of the app sandbox on Android using a known vulnerability.
We consider three attack vectors.
The first attack vector is through the installation of a malicious app on the device.
Android devices support the installation of apps through a number of app marketplaces, email attachments, URLs and via the Android Debug Bridge (ADB).
By default, many Android devices with Google Play installed will prevent the installation of apps from other sources.

Controlling potential sources of app installation is a good security feature because app marketplaces provide additional protection to prevent the listing of malicious apps.
For example, the Google Play marketplace uses Bouncer to automatically analyse apps as well as removing any apps which are reported as malicious.
Nevertheless, Google Play is not immune: in 2012 0.02\% of apps were malicious~\cite{Zhou2012a}.
Alternative markets are also popular, particularly in countries such as China where the Google Play is not available. 
In 2012 between 0.20\% and 0.47\% of apps on alternative markets were malicious~\cite{Zhou2012a}.

The second attack vector is to persuade an existing app to download and execute code at runtime.
The most direct method available to an attacker is to upload to a marketplace a seemingly innocent app which contains support for dynamic code loading from a remote server under control of the attacker.
Neither static nor dynamic analysis of this app by the marketplace will uncover any malicious code, since it does exist in the app binary.
Instead the attacker can choose to send a malicious payload directly to the app for execution at a later point in time.
Note that dynamic code loading is not actually needed for this to work -- it just makes the task easier -- since previous work has shown that even on a platform such as iOS, which does not permit dynamic code loading, a Return oriented Programming (ROP)-based attack is relatively easy if the attacker creates an app with carefully crafted flaws~\cite{Wang2013a}.

The third attack vector is that the attacker injects malicious code directly into existing apps or programs already installed on the handset. 
For example, the addJavascriptInterface vulnerability (CVE-2012-6636) allows Javascript running in an Android WebView to execute arbitrary code in the same process address space as the vulnerable app.
This attack works when: (1) the WebView has the Javascript-to-Java bridge enabled; and (2) the attacker is able to inject a malicious Javascript payload into the network traffic between the phone and the web server. 
Many apps are vulnerable to this type of attack because ad-libraries typically embed adverts into apps by displaying HTML and Javascript inside a WebView with the Javascript-to-Java bridge enabled. 
The fix for the addJavascriptInterface vulnerability breaks backwards compatibility, and therefore apps are only immune to this attack if their target API level is greater than 16 and they are running on an Android device with an OS version of 4.2 or greater.
Our analysis shows that on \daGPAPISeventeenLaterDate, \daGPAPISeventeenEarlierProportion\ of handsets connecting to the Google Play store were still vulnerable to this attack.

@@

We require two sources of data for this study: (1) information on the distribution of installed versions of Android over time and (2) information on the critical vulnerabilities found to affect specific versions of Android.
These two datasets can then be combined to determine the proportion of handsets at risk of attack from specific vulnerabilities for particular periods of time.

Where possible we indicate the uncertainty in our results by presenting them $\pm$ one standard deviation and give results to 3 significant figures, this occasionally results in `$\pm\, 0$' when the standard deviation is small.\todolater{Do we want to use the 95 percentile instead}
This technique cannot account for systematic errors which we explore in \S\ref{sec:representative}.

@@

\subsection{Analysis 1: vulnerability of Android devices}\label{sec:exp:versionsecurity}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figures/proportioninsecure}
\caption{Proportion of devices running insecure, maybe secure and secure versions of Android against time.
The red vertical lines are caused by vulnerabilities being discovered with those which have the biggest impact annotated.
}
\label{fig:proportioninsecure}
\end{figure}
In this analysis we calculate the proportion of Android devices which are exposed to critical vulnerabilities and and discuss how that has changed over time.
Figure~\ref{fig:proportioninsecure} describes the proportion of Android devices susceptible to at least one critical vulnerability.

@@

\subsubsection{Limitations}
We still do not know where new versions of Android come from.
Over time newer versions of Android come to dominate, but quantifying the proportion due to new phone purchases verses the proportion due to updates of existing devices is unknown.
Device Analyzer is the only available large data set we know of with longitudinal traces of Android OS version strings.
Unfortunately this data set only records \daNumUpdatesUpgrades\ upgrades which, broken down over \daOSMonthsOfData\ months and \daNumSigOSVersions\ OS versions is only \daUpdatesPerMonthPerVersion\ updates per version per month. This is not enough to build a statistically significant prediction of the expected transition to newer versions of Android.

>>>>

\section{Introduction}

Modern mobile operating systems provide several layers of security whose combined aim is to prevent app malware from taking control of a device. 
In this paper we explore the security layers present in the Android ecosystem and determine which layers are effective and which are not.

Our focus is on malware whose aim is to gain full control of the device rather than merely abuse permissions granted to it by the user.
In this context, by full control, or \emph{root}, on a device we mean the malware has unfettered access to all data stored on the device and can control all network interfaces in the device with the same flexibility as the operating system itself.
We have selected our focus for two reasons: no previous published work has quantified how vulnerable Android devices are to malware gaining root, and secondly the negative impact on the device owner is much higher for malware which gains root than malware which does not: malware with root can not only perform arbitrary actions on the device, but it is also impossible for the device owner to remove the malware reliably since even a factory reset does not guarantee that the malware has been successfully removed.

This is an important problem: previous studies have shown that a significant proportion of malware attempts to gain root.
For example, in 2012 between 36.7\%~\cite{Zhou2012b} and 40\%~\cite{Zhou2012a} of malware for Android contained exploits designed to gain root-level access.

Android users are protected from malicious apps by up to three levels of defence. 
The first level of defence comes from an app marketplace such as Google Play. 
A marketplace can use three mechanisms to prevent malicious apps from appearing in the marketplace: firstly by economic disincentives, including charging developers to create an account and cancelling accounts used to upload malicious apps; secondly by technical means, such as static and dynamic analysis on submitted apps; and finally via social feedback between users, including the provision of user reviews and ratings.

The second line of defence occurs on the handset at installation time, and includes checking the signature of any app installation or update as well as verifying the app by local anti-virus tools (e.g. Google's app verification feature). 
Finally, the third line of defence takes place at runtime: Android apps are run in a \emph{sandbox} which prevents the app from unauthorised actions such as accessing any permissions which weren't granted by the user at install time; in addition, anti-virus products may scan the list of installed apps periodically to detect malware.

Sandboxes only provide protection if they are free from design errors and implementation flaws. 
Both of these have occurred in Android, as indeed happens today in many large security systems relying on the sandbox model. 
Maintaining sandbox security therefore relies on providing timely updates to fix design and implementation flaws as and when they are discovered.

In this paper we show that the Android sandbox only provides effective security on \daMeanSecurityPercTwosfNominal\ of devices today.
\footnote{In contrast to Google's claims that 100\% of Android devices are protected by sandboxes and permissions~\cite{Patterson2013}}
This is because most devices do not receive timely updates. 
Therefore the security of the Android ecosystem is currently reliant on the defences provided by the app marketplace and (when enabled) anti-virus systems. 
This insight allows us to provide users with a list of concrete actions to improve device security. 
It also raises broader questions about how modern mobile device security is most effectively achieved, how a secure ecosystem should be construct and managed, and whether security by obscurity is the right way forward.

In summary, our contributions are:
\begin{itemize}
 \item We quantify the Android update ecosystem showing how the OS version distribution changes and the frequency of updates.
 \item We created an open database to collate quantitative information on vulnerabilities which affect Android and use this to determine the proportion of devices they affect and how that changes over time.
 \item We demonstrate that the security of Android relies on the centralisation of app deployment via a marketplace, and the use of install-time anti-virus checks rather than through the security of the app sandbox.
 \item We use our analysis to develop advice for Android users on what steps they should take to increase the security of their device.

\end{itemize}

\section{Threat model and attack vectors}
\label{sec:threatmodel}

In this paper we assume the defender can deploy security controls at three levels: in an online marketplace, at app installation, and during app execution.
We then consider three attack vectors an attacker can use to attempt to gain root on an Android handset.

The first attack vector is through the installation of a malicious app on the device.
Android devices support the installation of apps through a number of app marketplaces, email attachments, URLs and via the Android Debug Bridge (ADB).
By default, many Android devices with Google Play installed will prevent the installation of apps from other sources, and therefore an attacker needs to compromise all three levels of security.
The Google Play marketplace uses Bouncer to automatically analyse apps as well as removing any apps which are reported as malicious.
Nevertheless, Google Play is not immune: in 0.02\% of apps were malicious~\cite{Zhou2012a}.
Alternative markets are also popular, particularly in countries such as China where the Google Play is not available. 
In 2012 between 0.20\% and 0.47\% of apps on alternative markets were malicious~\cite{Zhou2012a}.

The second attack vector is to persuade an existing app to download and execute code at runtime.
The most direct method available to an attacker is to upload to a marketplace a seemingly innocent app which contains support for dynamic code loading from a remote server under control of the attacker.
Neither static nor dynamic analysis of this app by the marketplace will uncover any malicious code, since it does not exist in the app binary.
The attacker can choose to send a malicious payload directly to the app for execution at a later point in time.
The marketplace can detect explicit use of dynamic code loading, even if malicious code cannot be detected. Note that dynamic code loading is not actually needed for this to work however -- it just makes the task easier -- since previous work has shown that even on a platform such as iOS, which does not permit dynamic code loading, a Return-Oriented Programming (ROP)-based attack is relatively easy if the attacker creates an app with carefully crafted flaws~\cite{Wang2013a}.

The third attack vector is that the attacker injects malicious code directly into existing apps or programs already installed on the handset. 
For example, the addJavascriptInterface vulnerability (CVE-2012-6636) allows JavaScript running in an Android WebView to execute arbitrary code in the same process address space as the vulnerable app.
This attack works when: (1) the WebView has the JavaScript-to-Java bridge enabled; and (2) the attacker is able to inject a malicious JavaScript payload into the network traffic between the phone and the web server. 
Many apps are vulnerable to this type of attack because ad-libraries typically embed adverts into apps by displaying HTML and JavaScript inside a WebView with the JavaScript-to-Java bridge enabled. 
The fix for the addJavascriptInterface vulnerability breaks backwards compatibility, and therefore apps are only immune to this attack if their target API level is greater than 16 and they are running on an Android device with an OS version of 4.2 or greater.
Our analysis shows that as of \daGPAPISeventeenLaterDate, \daGPAPISeventeenEarlierProportion\ of handsets connecting to the Google Play store were still vulnerable to this attack.
\todo{Do we want to talk about the dhcp vulnerability here?}

In summary, the best place to prevent attacks is at runtime, since all three attack vectors can be prevented at this level. 
Unfortunately, as we will demonstrate, the sandbox for Android apps is weak and anti-virus is either not installed or not activated by default.
Therefore many Android users implicitly rely on the marketplace for protection by removing apps which are detected as malicious. This detection can only be derived from static and dynamic analysis or after reports of malicious behaviour from users once it is made available for download.

@@

We require two sources of data to determine whether the Android sandbox is secure or not: (1) information on the distribution of installed versions of Android over time and (2) information on the critical vulnerabilities found to affect specific versions of Android.
These two datasets can then be combined to determine the proportion of sandboxes at risk of attack from specific vulnerabilities for particular periods of time.

We indicate the uncertainty in our results by presenting them $\pm$ one standard deviation and give results to 3 significant figures, this occasionally results in `$\pm\, 0$' when the standard deviation is small.
\todolater{Do we want to use the 95 percentile instead}
This technique cannot account for systematic errors which we explore in \S\ref{sec:representative}.

@@

\subsection{Analysis 1: vulnerability of Android devices}\label{sec:exp:versionsecurity}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figures/proportioninsecure}
\caption{Proportion of devices running insecure, maybe secure and secure versions of Android against time.
The red vertical lines are caused by vulnerabilities being discovered with those which have the biggest impact annotated.
}
\label{fig:proportioninsecure}
\end{figure}
In this analysis we calculate the proportion of Android devices which are exposed to critical vulnerabilities and discuss how that has changed over time.
Figure~\ref{fig:proportioninsecure} describes the proportion of Android devices susceptible to at least one critical vulnerability.

@@

\subsubsection{Limitations}
We still do not know where new versions of Android come from.
Over time newer versions of Android come to dominate, but quantifying the proportion due to new phone purchases verses the proportion due to updates of existing devices is unknown.
Device Analyzer is the only available large data set we know of with longitudinal traces of Android OS version strings.
Unfortunately this data set only records \daNumUpdatesUpgrades\ upgrades which, broken down over \daOSMonthsOfData\ months and \daNumSigOSVersions\ OS versions is only \daUpdatesPerMonthPerVersion\ updates per version per month. This is not enough to build a statistically significant prediction of the expected transition to newer versions of Android.
If manufacturers shipped security updates without changing the build number then we would not be able to detect this, but that would also indicate a level of incompetence likely to damage security and so can be discounted.
